### 2025-11-09 thesis/implementation/index.tex
Finish impl section

Frequent CUDA out-of-memory (OOM) errors made the evaluation interpretation non
trivial as accuracy scores became inaccurate due to pipeline failures. This persistet as a

second development challenge throughout the implementation phase. These memory exhaustion issues
occurred rather unpredictable during both development and benchmarking,
depending on the database schema and question asked. As \textsc{Natural} uses a
code representation prompt, the database schema is inlined into the prompt
during inference. If \textsc{Natural} is executed against large databases, even
with schema subsetting, the context window can be exceeded or the KV-Cache can
exhaust the rest of the available GPU memory (typically around 2-4 gigabytes)
after models have been loaded. This required extensive trial-and-error
debugging to identify the root causes and implement workarounds.

@@

The decision to implement \textsc{Natural} in Rust rather than Python
represents perhaps the most impactful architectural choice made during the
implementation phase. Choosing Rust presented the opportunity to port the
research system into production environments soon after, but came at the cost
of having to port existing benchmarking infrastructure and not being able to
exactly reuse the inference code used by \textsc{OmniSQL}. This decision
significantly increased implementation time, likely doubling the development
effort compared to what would have been required in Python.

Using Python would have provided immediate access to existing machine
learning infrastructure and could have leveraged existing implementations by
previous researchers directly in the implementation of \textsc{Natural}.

The rather mature Python ecosystem for NLP and ML would have eliminated the
need to develop abstractions for model inference, vector database operations,
and benchmark evaluation.

@@

>>>>

Frequent CUDA out-of-memory (OOM) errors made the evaluation interpretation non
trivial as accuracy scores became inaccurate due to pipeline failures. This
persistet as a second development challenge throughout the implementation
phase. These memory exhaustion issues occurred rather unpredictable during both
development and benchmarking, depending on the database schema and question
asked. As \textsc{Natural} uses a code representation prompt, the database
schema is inlined into the prompt during inference. If \textsc{Natural} is
executed against large databases, even with schema subsetting, the context
window can be exceeded or the KV-Cache can exhaust the rest of the available
GPU memory (typically around 2-4 gigabytes) after models have been loaded. This
required extensive trial-and-error debugging to identify the root causes and
implement workarounds.

@@

The decision to implement \textsc{Natural} in Rust rather than Python
represents perhaps the most impactful architectural choice made during the
implementation phase. Choosing Rust presented the opportunity to port the
research system into production environments soon after, but came at the cost
of having to port existing benchmarking infrastructure and not being able to
exactly reuse the inference code used by \textsc{OmniSQL}. This decision
significantly increased implementation time, likely doubling the development
effort compared to what would have been required in Python. Using Python would
have provided immediate access to existing machine learning infrastructure and
could have leveraged existing implementations by previous researchers directly
in the implementation of \textsc{Natural}. The rather mature Python ecosystem
for NLP and ML would have eliminated the need to develop abstractions for model
inference, vector database operations, and benchmark evaluation.

@@

\subsubsection{Lessons Learned and Future Recommendations}

The implementation of \textsc{Natural} showed multiple practical challenges
associated with researching and implementing NL2SQL systems. The approach used
in this thesis highlights several areas of consideration for future reserch
projects in this field.

\paragraph{Hardware Infrastructure Requirements}

Research of modern NL2SQL systems induces steep hardware requirements onto
researchers. Priority should be given to securing access to significantly
powerful hardware and computing infrastructure for faster iteration cycles.

While the 24GB VRAM constraint of the RTX 3090, may be reasonable for most
consumer grade applications, it proved insufficient for efficient iteration in
this field, especially given the involement of large language models.
Systems with 48GB+ of VRAM would have enabled the usage of larger, more capable
models without the quantization compromises that may impact system accuracy.
More modern hardware (even on the consumer grade level) would allow for a
tremendous benchmarking speed up (eg, using a 5090 would have resulted in
nearly 200\% increase in speed).

More critically, the availability of multiple GPUs would have enabled parallel
experimentation and benchmarking, which could transform multi-day iteration
cycles into more manageable timeframes. The ability to run concurrent
experiments with different configurations, models, or algorithmic approaches
would dramatically accelerate the research and development process and enable
more thorough experimental validation.

\paragraph{Technology Stack}

The implications of choosing between Rust and Python underlines the importance
of technology choices with research goals and time constraints. For pure
research projects focused on algorithmic development and experimentation,
Python's ecosystem and extensive existing implementations provide clear
advantages that likely outweigh the deployment benefits of compiled languages
like Rust.

However, for projects with a clear path to production deployments or those
intended to bridge academic research with real-world applications, the
additional development overhead of systems programming languages like Rust can
be justified. Overall, hardware availability weighs significantly more than
language choices when it comes to the implementation timeline.

Future implementations in this field should likely consider a two-staged
approach: prototyping and algorithm validation in Python to leverage existing
tools and accelerate iteration, followed by a production port in a systems
language like Rust once the algorithmic approach is validated and stable.

\paragraph{Development Methodology Recommendations}

The challenges encountered suggest several methodological improvements for
future projects:

Implementing benchmarking and validation infrastructure early in the
development process is crucial when doing algorithmic development to gain an
understanding of the relative improvement or setback of different approaches.

Establishing performance baselines and using incremental validation procedures
can help ensure that development effort is focused on improvements that in fact
improve performance.

The long running benchmarks made it difficult to validate whether
individual changes were beneficial, leading to uncertainty about the
effectiveness of specific optimizations. In situations with highly constrained
compute resources, a representative subset of prevalent benchmarking datasets
like \textsc{Spider} and \textsc{Bird} may be defined to determine the relative
improvement between approaches before executing a full benchmark run.

Finally, maintaining detailed logs and metrics throughout the development
process showed to be crucial when unexpected errors occur. Identifying root
causes of pipeline failures was rather trivial due to having an advanced
benchmarking setup with log collection.

\paragraph{Research Impact and Production Readiness}

Despite all challenges encountered in the implementation phase, the Rust
implementation of \textsc{Natural} demonstrate that it is possible to port
research algorithms from academic environments into production-ready systems.
The resulting implementation provides a strong foundation for real-world
deployments in database environments, which would be significantly more
challenging with a Python-based research prototype.

This production readiness ensures that all algorithmic contributions can have
practical impact beyond pure academic evaluation, potentially influencing how
NL2SQL systems are deployed in real database environments. A simple
recommendation between research velocity and production readiness can not be
given as it ultimately depends on the intended impact and longevity of the
research contributions and implementation.
